---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Cailey Fay"
date: "2/15/26"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
staff <- read_csv("data/instructional-staff.csv")
```

I think when we convert to long-form data, we will have 6 columns. If there are five different faculty types and 11 years, I think that means we will have 5x11 = 55 rows? 

To get the data long:
```{r}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
staff_long
```
This confirms my thought that there would be 55! 

To try to plot this as a line graph with years on the x, the percentage of hires on the y, and 5 lines for the diferent faculty types:
```{r}
staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()
```
This doesn't quite work. 

### Exercise 1

```{r}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  labs(title = "Percentage of Faculty Types Hired by Year",
       x= "Year",
       y= "Percentage of Hires",
       color = "Faculty Type")

```

### Exercise 2

Suppose the objective of this plot was to show that the proportion of part-time faculty have gone up over time compared to other instructional staff types. What changes would you propose making to this plot to tell this story?

I would make all of the lines grey except the part time faculty. I think this pretty clearly communicates that the percentage of part time faculty hires is increasing. 
```{r}

staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  labs(title = "Percentage of Faculty Types Hired by Year",
       x= "Year",
       y= "Percentage of Hires",
       color = "Faculty Type") +
    scale_color_manual(values = c(
    "Full-Time Tenured Faculty" = "grey",
    "Full-Time Tenure-Track Faculty" = "grey",
    "Full-Time Non-Tenure-Track Faculty" = "gray",
    "Part-Time Faculty" = "red",
    "Graduate Student Employees" = "grey"
  )) 


```

### Exercise 3

...

Add exercise headings as needed.
